### 2012-09-27 doc/prelim/document/chapters/nonlinear_problem/nonlinear_problem.tex
starting work on the research proposal

\chapter{Stochastic Solution Methods for Nonlinear Problems}
\label{ch:nonlinear_problem}
Nonlinear equation sets are a common occurence in multiphysics
problems. Systems of partial differntial equations such as those that
describe fluid flow or more general transport processes when
discretized by conventional methods yield discrete sets of stiff
equations with nonlinearities present in the variables. Traditionally,
such systems have been solved by linearizing them such that the
nonlinearities in the variable are eliminated and more traditional
linear methods can be used for solutions. Often characterized as
segregated methods where physics operators are split and their action
on the system approximated in steps, such methods lack consistency and
accuracy in resolving the nonlinear component of the solution. In the
last 30 years, fully consistent nonlinear methods based on Newton's
method have become more popular and many advances have been made in
the computational physics field to employ these methods.

Currently, Krylov methods are used with Newton's method to form a
robust solution sequence for these types of problems. However, given
our goals of advancing stochastic solver technology for linear
problems, we aim to apply them to these fully consistent schemes in
order to gauge if some improvements can be made. We therefore devise a
new method using MCSA in conjuction with new methods, accomodating the
fact that MCSA requires an explicitly formed linear operator in its
implementation as compared to Krylov methods. In addition to handling
this dependency, we will construct a parallel strategy for a nonlinear
MCSA method.

@@

\section{Inexact Newton Methods}
\label{sec:newton_methods}
Inexact Newton methods arise when the Jacobian operator is not exactly
inverted, resulting in an inexact Newton correction as initially
described by Dembo and others \citep{dembo_inexact_1982}. For common
sparse nonlinear systems, which in turn yield a sparse Jacobian
matrix, this situation occurs when conventional iterative methods are
applied. In their definition, Dembo formulated inexact methods such
that they are independent of the linear method used to solve for the
Newton correction and therefore are amenable to use with any linear
solver. Furthermore, they bind the convergence of the outer nonlinear
iteration to the inner linear iteration such that:
\begin{equation}
  ||\ve{J}(\ve{u}^k)\delta \ve{u}^k + \ve{F}(\ve{u}^k)|| \leq \eta^k
  ||\ve{F}(\ve{u}^k)||\:,
  \label{eq:inexact_newton_forcing}
\end{equation}
where $\eta^k \in [0,1)$ is defined as the \textit{forcing term} at
  the $k$ iterate. Eq~(\ref{eq:inexact_newton_forcing}) then states
  that the residual generated by the linear solver is bound by the
  nonlinear residual and how tightly it is bound is defined by the
  forcing term. This is useful in that we can vary how tightly coupled
  the convergence of the linear iterations used to generate the Newton
  correction is to the nonlinear iteration by relaxing or tightening
  the convergence properties on the linear iterative method. As a
  result, strategies for determining the forcing term can vary
  depending on the problem type and can greatly affect the convergence
  of the method or even prohibit convergence
  \citep{eisenstat_choosing_1996}. In addition, globalization methods
  may be used to modify the Newton correction in a more desireable
  direction such that convergence properties can be improved when the
  initial guess for $\ve{u}$ is poor
  \citep{pawlowski_globalization_2006}.

@@

With many Krylov methods available, which to use with the Newton
method is dependent on many factors including convergence rates and
memory usage. Several studies have been performed to investigate this
\citep{mchugh_inexact_1993,knoll_newton-krylov_1995}. In their
numerical studies in 1995, Knoll and McHugh used the set of highly
nonlinear and stiff convection-diffusion-reaction equations to solve a
set of tokamak plasma problems with the goal of measuring solver
performance with Newton's method. They note several tradeoffs in using
Krylov methods with the Newton solver. The first is that the
optimization condition that results from the constraints (e.g. the
minimization of the GMRES residual over the Krylov space) can be
relaxed by restricting the size of the subspace such that only a fixed
number of subspace vectors may be maintained, thus reducing memory
requirements. We can also relax the optimization condition by instead
restarting the recurrence relation with a new set of vectors once a
certain number of vectors have been generated, the optimization
condition is maintain over that particular set of vectors, however,
Knoll and McHugh note that this ultimately slows the convergence rate
as compared to keeping all vectors as the new set of vectors is not
necessarily orthogonal to the previous set, and therfore not optimal
over the entire iteration procedure. The orthogonality condition can
be relaxed by using a recurrence relation that does not generate a
strictly orthonormal basis for the Krylov subspace such as the Lanzcos
biorthogonalization procedure, resulting in memory savings due to the
shorter Lanzcos recurrence relation.

@@

\subsubsection{Automatic Differentiation for Jacobian Generation}
\label{subsubsec:automatic_differentiation}
If it is acceptable to store the actual Jacobian matrix, other methods
are available to construct it without requiring hand-coding and
evaluating derivatives, thus eliminating the associated issues. In
addition, if any additional equations are added to the system or a
higher order functional approximation is desired, it would be useful
to avoid regenerating and coding these derivatives. Becoming more
prominent in the 1990's, \textit{automatic differentiation} is a
mechanism by which the derivatives of a function can be generated
automatically by evaluating it. Automatic differntiation is built on
the concept that all functions discretely represented in a computer
are ultimately represented by elementary mathematical operations. If
the chain rule is applied to those elementary operations, then the
derivatives of those functions can be computed to the order of
accuracy of their original discretization in a completely automated
way \citep{averick_computing_1994}.

@@

Of interest to Bartlett, Averick, and the many others that have
researched automatic differentation are measuring its performance
relative to hand-coded derivatives and capturing the Jacboian matrix
from matrix-free approximations. Given their element-based function
evaluation scheme, Bartlett's work varied the number of degrees of
freedom per element and compared both the floating point operation
count and CPU time for both the templated automatic differentiation
method and hand-coded derivatives for Jacobian evaluations. Although
they observed a 50\% increase in floating point operations in the
templated method over the hand-coded method, run times were observed
to be over 3 times faster for the templated method. They hypothesize
that this is due to the fact that the element-based formulation of the
templated method is causing better utilization of cache and therefore
faster data access. Furthermore, they observed linear scaling behavior
for automatic differentiation as the number of degrees of freedom per
element were increased up to a few hundred. Based on these results,
this type of automatic differentation formulation was deemed
acceptable for use in large-scale production physics codes.

\section{The FANM Method}
\label{sec:fanm}
In production physics codes based on nonlinear equations sets,
Newton-Krylov methods are the primary means of generating a fully
consistent solution scheme
\citep{evans_development_2006,evans_enhanced_2007,gaston_parallel_2009,godoy_parallel_2012}. Typically,
for large scale simulations these problems are memory limited due to
the subspaces generated by robust Krylov methods. Often a matrix-free
approach is chosen to relax memory requirements over directly
generating the Jacobian matrix. However, as we observed in previous
sections, these matrix-free methods suffer from poorly scaled problems
and the first order error introduced by the Jacobian approximation. In
addition, it was observed that eventually, the savings induced by the
matrix-free approach is eventually amortized over a number of
nonlinear iterations where it becomes more efficient computationally
to instead form the Jacobian.

In Chapter~\ref{ch:stochastic_methods}, we focused our efforts on
developing and improving Monte Carlo methods for inverting linear
systems. These methods, when used to accelerate a stationary method in
MCSA, enjoy a robust implementation and exponential convergence
rates. Further, the only storage required is that of the full linear
system including the linear operator so that we may generate
transition probabilities for the random walk sequence. Although this
requires more storage to represent the linear system than that of a
Krylov method where the operator is not required, we do not incur any
additional storage costs once the iteration sequence
begins. Furthermore, in the context of nonlinear problems, the
Jacobian matrix that we are required to generate for the Monte Carlo
solvers may be generated at will from the nonlinear functions in the
Newton system using automatic differentation. Not only do we then have
a simple and automated way to generate the Jacobian, but we also enjoy
a Jacobian of numerical precision equivalent to that of our function
evaluations. We therefore propose the \textit{Forward-Automated
  Newton-MCSA} (FANM) method that utilizes all of the above
components. We hypothesize that such a method will be competetive with
Newton-Krylov methods not only from a convergence and timing
perspective, but also relax scaling requirements of matrix-free
methods and memory costs of both matrix-free and fully formed Jacobian
methods to allow the application developer to solve problems of finer
discretization and higher order functional representations while
maintaining an robust an efficient parallel implementation.

@@

In his discussion on orthogonalization schemes for Krylov subspaces,
he provides us with space and time complexities for these
operations. To find a solution vector $\ve{x} \in \mathbb{R}^{N \times
  N}$ with $m$ Krylov iterations (and therefore $m+1$ subspace
vectors) the storage requirement is then $(m+1) N$ for most common
orthogonalization techniques. Using our simple example above (even
though we did not see any gains in storage over the full matrix
representation), if it takes 10 GMRES iterations using Arnoldi
orthogonalization to converge to a solution, then a total of 66 pieces
of data are required to be stored instead of 36 for the explicitly
formed matrix case. For larger sparse matrices, this disparity in
memory requirements will be even greater, especially for
ill-conditioned systems that require many GMRES iterations to
converge.

\subsection{Parallel FANM Implementation}
\label{subsec:parallel_fanm}
Like the parallelization of the MCSA method described in
\S~\ref{sec:parallel_stochastic_methods}, a parallel FANM method
relies on a basic set of parallel matrix-vector operations as well as
the global residual and Jacobian assembly procedure described in
\S~\ref{subsubsec:automatic_differentiation}. Consider the Newton
iteration scheme in Eq~(\ref{eq:newton_iteration}). We must first
assemble the linear system in parallel through the element-wise
function evaluations to generate both the global Jacobian operator and
the global residual vector on the right hand side. Per Bartlett's
work, efficient and automated parallel mechanisms are available to do
this through a sequence of scatter/gather operations. With these tools
available for residual and Jacobian generation, the remainder of the
parallel procedure is simple, with the linear Newton correction system
solved using the parallel MCSA method as previously described and the
Newton correction applied to the previous iterate's solution through a
vector update operation.

As Newton methods are formulated independent of the inner linear
solver generating the Newton corrections, the actual performance of
the nonlinear iterations using MCSA is expected to be similar to that
of traditional Newton-Krylov methods. Furthermore, we expect to
achieve numerically identical answers with a Newton-MCSA method as
other Newton methods and we should indeed verify this. The parallel
performance of such a method will inherently be bound to the parallel
Monte Carlo implementation of the linear solver as the parallel
operations at the nonlinear iteration level are identical to those
that you would perform with a Newton-Krylov method. Matrix-free
formulation will have different parallel performance than these
methods, and therefore we should compare FANM perfomance to JFNK-based
schemes. More important than performance in the situation is the
reduced memory pressure that a FANM implementation provides, as
discussed in \S~\ref{subsec:fanm_storage}, because a FANM method will
not generate a subspace in the linear solver and compressed storage
for sparse matrices are utilized, we expect significant memory savings
over Newton-Krylov methods. We must measure the memory utilization of
both of these methods in order to quantify their differences and
provide additional analysis of the FANM method's merits, or lack
thereof.

>>>>

\chapter{Stochastic Solution Methods for Nonlinear Problems}
\label{ch:nonlinear_problem}
Nonlinear equation sets are a common occurence in multiphysics
problems. Systems of partial differntial equations such as those that
describe fluid flow or more general transport processes when
discretized by conventional methods yield discrete sets of stiff
equations with nonlinearities present in the variables. Traditionally,
such systems have been solved by linearizing them such that the
nonlinearities in the variables are eliminated and more traditional
linear methods can be used for solutions. Often characterized as
segregated methods where physics operators are split and their action
on the system approximated in steps, such methods lack consistency and
accuracy in resolving the nonlinear component of the solution. In the
last 30 years, fully consistent nonlinear methods based on Newton's
method have become more popular and many advances have been made in
the computational physics field to employ these methods.

Currently, Krylov methods are used with Newton's method to form a
robust solution sequence for these types of problems. However, given
our goals of advancing stochastic solver technology for linear
problems, we aim to apply them to these fully consistent schemes in
order to gauge if some improvements can be made. We therefore devise a
new method using MCSA in conjuction with new methods, accomodating the
fact that MCSA requires an explicitly formed linear operator in its
implementation as compared to Krylov methods. In addition to handling
this dependency, we will construct a parallel strategy for a nonlinear
MCSA-based method.

@@

\section{Inexact Newton Methods}
\label{sec:newton_methods}
Inexact Newton methods arise when the Jacobian operator is not exactly
inverted, resulting in an inexact Newton correction as initially
described by Dembo and others \citep{dembo_inexact_1982}. For common
sparse nonlinear systems, which in turn yield a sparse Jacobian
matrix, this situation occurs when conventional iterative methods are
applied. In their definition, Dembo formulated inexact methods such
that they are independent of the linear method used to solve for the
Newton correction and therefore are amenable to use with any linear
solver. Furthermore, they bind the convergence of the outer nonlinear
iteration to the inner linear iteration such that:
\begin{equation}
  ||\ve{J}(\ve{u}^k)\delta \ve{u}^k + \ve{F}(\ve{u}^k)|| \leq \eta^k
  ||\ve{F}(\ve{u}^k)||\:,
  \label{eq:inexact_newton_forcing}
\end{equation}
where $\eta^k \in [0,1)$ is defined as the \textit{forcing term} at
  the $k$ iterate. Eq~(\ref{eq:inexact_newton_forcing}) then states
  that the residual generated by the linear solver is bound by the
  nonlinear residual and how tightly it is bound is defined by the
  forcing term. This is useful in that we can vary how tightly coupled
  the convergence of the linear iterations used to generate the Newton
  correction is to the nonlinear iteration by relaxing or tightening
  the convergence properties on the linear iterative method. As a
  result, strategies for determining the forcing term can vary
  depending on the problem type and can greatly affect the convergence
  of the method or even prohibit convergence
  \citep{eisenstat_choosing_1996}. In addition, \textit{globalization
    methods} may be used to modify the Newton correction in a more
  desireable direction such that convergence properties can be
  improved when the initial guess for $\ve{u}$ is poor
  \citep{pawlowski_globalization_2006}.

@@

With many Krylov methods available, which to use with the Newton
method is dependent on many factors including convergence rates and
memory usage. Several studies have been performed to investigate this
\citep{mchugh_inexact_1993,knoll_newton-krylov_1995}. In their
numerical studies in 1995, Knoll and McHugh used the set of highly
nonlinear and stiff convection-diffusion-reaction equations to solve a
set of tokamak plasma problems with the goal of measuring solver
performance with Newton's method. They note several tradeoffs in using
Krylov methods with the Newton solver. The first is that the
optimization condition that results from the constraints (e.g. the
minimization of the GMRES residual over the Krylov space) can be
relaxed by restricting the size of the subspace such that only a fixed
number of subspace vectors may be maintained, thus reducing memory
requirements. We can also relax the optimization condition by instead
restarting the recurrence relation with a new set of vectors once a
certain number of vectors have been generated. The optimization
condition is maintain over that particular set of vectors, however,
Knoll and McHugh note that this ultimately slows the convergence rate
as compared to keeping all vectors as the new set of vectors is not
necessarily orthogonal to the previous set, and therfore not optimal
over the entire iteration procedure. The orthogonality condition can
be relaxed by using a recurrence relation that does not generate a
strictly orthonormal basis for the Krylov subspace such as the Lanzcos
biorthogonalization procedure, resulting in memory savings due to the
shorter Lanzcos recurrence relation.

@@

\subsubsection{Automatic Differentiation for Jacobian Generation}
\label{subsubsec:automatic_differentiation}
If it is acceptable to store the actual Jacobian matrix, other methods
are available to construct it without requiring hand-coding and
evaluating derivatives, thus eliminating the associated issues. In
addition, if any additional equations are added to the system or a
higher order functional approximation is desired, it would be useful
to avoid regenerating and coding these derivatives. Becoming more
prominent in the 1990's, \textit{automatic differentiation} is a
mechanism by which the derivatives of a function can be generated
automatically by evaluating it. Automatic differentiation is built on
the concept that all functions discretely represented in a computer
are ultimately represented by elementary mathematical operations. If
the chain rule is applied to those elementary operations, then the
derivatives of those functions can be computed to the order of
accuracy of their original discretization in a completely automated
way \citep{averick_computing_1994}.

@@

Of interest to Bartlett, Averick, and the many others that have
researched automatic differentation are measures of its performance
relative to hand-coded derivatives and capturing the Jacboian matrix
from matrix-free approximations. Given their element-based function
evaluation scheme, Bartlett's work varied the number of degrees of
freedom per element and compared both the floating point operation
count and CPU time for both the templated automatic differentiation
method and hand-coded derivatives for Jacobian evaluations. Although
they observed a 50\% increase in floating point operations in the
templated method over the hand-coded method, run times were observed
to be over 3 times faster for the templated method. They hypothesize
that this is due to the fact that the element-based formulation of the
templated method is causing better utilization of cache and therefore
faster data access. Furthermore, they observed linear scaling behavior
for automatic differentiation as the number of degrees of freedom per
element were increased up to a few hundred. Based on these results,
this type of automatic differentation formulation was deemed
acceptable for use in large-scale, production physics codes.

\section{The FANM Method}
\label{sec:fanm}
In production physics codes based on nonlinear equations sets,
Newton-Krylov methods are the primary means of generating a fully
consistent solution scheme
\citep{evans_development_2006,evans_enhanced_2007,gaston_parallel_2009,godoy_parallel_2012}. Typically,
for large scale simulations these problems are memory limited due to
the subspaces generated by robust Krylov methods. Often, a matrix-free
approach is chosen to relax memory requirements over directly
generating the Jacobian matrix and facilitate the
implementation. However, as we observed in previous sections, these
matrix-free methods suffer from poorly scaled problems and the first
order error introduced by the Jacobian approximation. In addition, it
was observed that the savings induced by the matrix-free approach is
eventually amortized over a number of nonlinear iterations where it
becomes more efficient computationally to instead form the Jacobian.

In Chapter~\ref{ch:stochastic_methods}, we focused our efforts on
developing and improving Monte Carlo methods for inverting linear
systems. These methods, when used to accelerate a stationary method in
MCSA, enjoy a robust implementation and exponential convergence
rates. Further, the only storage required is that of the full linear
system including the linear operator so that we may generate
transition probabilities for the random walk sequence. Although this
requires more storage to represent the linear system than that of a
Krylov method where the operator is not required, we do not incur any
additional storage costs once the iteration sequence
begins. Furthermore, in the context of nonlinear problems, the
Jacobian matrix that we are required to generate for the Monte Carlo
solvers may be generated at will from the nonlinear functions in the
Newton system using automatic differentation. Not only do we then have
a simple and automated way to generate the Jacobian, but we also enjoy
a Jacobian of numerical precision equivalent to that of our function
evaluations. We therefore propose the \textit{Forward-Automated
  Newton-MCSA} (FANM) method that utilizes all of the above
components. We hypothesize that such a method will be competetive with
Newton-Krylov methods not only from a convergence and timing
perspective, but also relax scaling requirements of matrix-free
methods and memory costs of both matrix-free and fully formed Jacobian
methods to allow the application developer to solve problems of finer
discretization and higher-order functional representations while
maintaining a robust and efficient parallel implementation.

@@

In his discussion on orthogonalization schemes for Krylov subspaces,
Saad provides us with space and time complexities for these
operations. To find a solution vector $\ve{x} \in \mathbb{R}^{N \times
  N}$ with $m$ Krylov iterations (and therefore $m+1$ subspace
vectors) the storage requirement is then $(m+1) N$ for most common
orthogonalization techniques. Using our simple example above (even
though we did not see any gains in storage over the full matrix
representation), if it takes 10 GMRES iterations using Arnoldi
orthogonalization to converge to a solution, then a total of 66 pieces
of data are required to be stored instead of 36 for the explicitly
formed matrix case (we do not consider graph storage here for
comparison as both methods will require the graph for parallel
operations). For larger sparse matrices, this disparity in memory
requirements will be even greater, especially for ill-conditioned
systems that require many GMRES iterations to converge.

\subsection{Parallel FANM Implementation}
\label{subsec:parallel_fanm}
Like the parallelization of the MCSA method described in
\S~\ref{sec:parallel_stochastic_methods}, a parallel FANM method
relies on a basic set of parallel matrix-vector operations as well as
the global residual and Jacobian assembly procedure described in
\S~\ref{subsubsec:automatic_differentiation}. Consider the Newton
iteration scheme in Eq~(\ref{eq:newton_iteration}). We must first
assemble the linear system in parallel through the element-wise
function evaluations to generate both the global Jacobian operator and
the global residual vector on the right hand side. Per Bartlett's
work, efficient and automated parallel mechanisms are available to do
this through a sequence of scatter/gather operations. With these tools
available for residual and Jacobian generation, the remainder of the
parallel procedure is simple, with the linear Newton correction system
solved using the parallel MCSA method as previously described and the
Newton correction applied to the previous iterate's solution through a
parallel vector update operation.

As Newton methods are formulated independent of the inner linear
solver generating the Newton corrections, the actual performance of
the nonlinear iterations using MCSA is expected to be similar to that
of traditional Newton-Krylov methods. Furthermore, we expect to
achieve numerically identical answers with a Newton-MCSA method as
other Newton methods and we should indeed verify this. The parallel
performance of such a method will inherently be bound to the parallel
Monte Carlo implementation of the linear solver as the parallel
operations at the nonlinear iteration level are identical to those
that you would perform with a Newton-Krylov method. Matrix-free
formulations will have different parallel performance than these
methods, and therefore we should compare FANM perfomance to JFNK-based
schemes. More important than performance in the situation is the
reduced memory pressure that a FANM implementation provides, as
discussed in \S~\ref{subsec:fanm_storage}, because a FANM method will
not generate a subspace in the linear solver and compressed storage
for sparse matrices are utilized, we expect significant memory savings
over Newton-Krylov methods. We must measure the memory utilization of
both of these methods in order to quantify their differences and
provide additional analysis of the FANM method's merits, or lack
thereof.

### 2012-09-27 doc/prelim/document/chapters/research_proposal/research_proposal.tex
starting work on the research proposal

\chapter{Research Proposal}
\label{ch:research_proposal}

After all of the background information and explanations of the new
work, this section is the most important of all. Here I will clearly
define the questions I aim to answer and the strategy that I will use
to attempt to answer them. I think that it will be important to pick a
challenge problem here that FANM has a chance to show gains in either
memory footprint or parallelism over current methods. In particular, I
would like to have this challenge problem derive from nuclear reactor
multiphysics. I think Drekar would be a great test bed for FANM once
complete as it has all the tools to build a FANM solve (it is in fact
based on the same core libraries as Chimera) and a complex reactor
core assembly multiphysics system capable of scaling to large
machines. In addition, all of the complex diagnostic tools necessary
for analysis of FANM performance as compared to the other methods
should be available.

>>>>

\chapter{Research Proposal}
\label{ch:research_proposal}
To develop the parallel MCSA and FANM methods, a research plan is
proposed to drive their development and quantify their merits and
feasiblity as compared to more conventional methods for linear and
nonlinear problems. We seek to rigorously verify both of these methods
and there implementation and will describe the means by which that
will be accomplished. Further more, the numerical experiments
performed will be constructed within a software framework that will
permit an agile development strategy.
